Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-17 latex/main.tex
Updates from Overleaf

Before characterizing recombination events in detail,
we sought to remove probable artifactual recombinants
due to correlated bases
being treated as individual sites by the HMM.
Correlated bases can be caused by
systematic errors in genome sequences
associated with sequencing protocols \cite{Turakhia2020},
multi-nucleotide substitutions \cite{DeMaio2024},
genuine deletions, and
alignment errors (e.g., deletions due to MAFFT misalignment \cite{Hunt2024}).

We characterized artifactual recombination events
as having fewer than four such loci supporting the left or right parent
on either side of a suggested breakpoint.
We treated adjacent and near-adjacent sites
(separated by at most three bases) as single loci.
Also, when counting loci supporting a focal parent,
we penalized loci where the character(s)
in the recombinant matches that in the non-focal parent
but not in the focal parent.

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
Evidence that this filter effectively flagged artifacts can be seen in the
clear association between the AmpliSeq V1 kit and recombination events
that we identify as artifactual.
This primer scheme was used to sequence XX of the samples
associated with an artifactual recombination event,
in contrast to 15 of the ``robust'' events (Figure~\ref{fig:recombinant_qc}),
strongly implicating it as a common source of problematic data.

Three copying patterns in Figure~\ref{fig:copying_pattern_examples}
Q1, 686
Q3, 22029, Delta lineage-defining deletion (REF)
There are others associated with a known amplicon dropout caused by this particular deletion.
Q4, 22194
True MNS
Alignment error

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{recombinant_qc}
\caption{
Classification of ``robust'' and artifactual recombination events.
Scatterplot of recombination events by the net number of supporting loci
for the left parent (x-axis) and for the right parent (y-axis). The shaded
regions highlight artifactual recombination events, which have fewer than four
net supporting loci on one side or both sides of the suggested breakpoint.
Colors indicate the primer scheme used for sequencing. The pie charts show
breakdowns of the recombination events by primer scheme per quadrant.
Recombination events associated with the origins of Pango X lineages are labeled.
}
\label{fig:recombinant_qc}
\end{figure}

>>>>

Before characterizing recombination events in detail,
we sought to remove probable artifactual recombinants
due to correlated bases, which can be caused by
systematic errors in genome sequences
associated with sequencing protocols \cite{Turakhia2020},
multi-nucleotide substitutions (MNS) \cite{DeMaio2024}, and
alignment errors (e.g., deletions due to MAFFT misalignment \cite{Hunt2024}).

We characterized artifactual recombination events
as having fewer than four such loci supporting the left or right parent
on either side of a suggested breakpoint.
We treated adjacent and near-adjacent sites
(separated by at most three bases) as single loci.
Also, when counting loci supporting a focal parent,
we penalized loci where the character(s)
in the recombinant matches that in the non-focal parent
but not in the focal parent.
Illustrative copying patterns in Figure~\ref{fig:copying_pattern_examples}
show the evidence for some QC-failing recombinants:
(1) Q1, 686;
(2) Q3, MNS at 13423-13424 \cite{DeMaio2024};
(3) Q4, 22194 (possibly MAFFT alignment error \cite{Hunt2024}).

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
Evidence that this filter effectively flagged artifacts can be seen
in the clear association between the AmpliSeq V1 kit and
recombination events that we identified as artifactual.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{recombinant_qc}
\caption{
Quality control of recombination events.
Scatterplot of recombination events by the net number of supporting loci
for the left parent (x-axis) and for the right parent (y-axis). The shaded
regions highlight artifactual recombination events, which have fewer than four
net supporting loci on one side or both sides of the suggested breakpoint.
Colors indicate the primer scheme used for sequencing. The pie charts show
breakdowns of the recombination events by primer scheme per quadrant.
Recombination events associated with the origins of Pango X lineages are labeled.
}
\label{fig:recombinant_qc}
\end{figure}
